### 2019-10-31 Supplement_forShort.Rmd
revamping the manuscript for re-submission, making shortened version and separate supplement for stuff removed from the main manuscript

# The Evolution of Siphonophore Tentilla as Specialized Tools for Prey Capture {-}

Alejandro Damian-Serrano^1^^,‡^, Steven H.D. Haddock^2^, Casey W. Dunn^1^

^1^ Yale University, Department of Ecology and Evolutionary Biology, 165 Prospect St., New Haven, CT 06520, USA

^2^ Monterey Bay Aquarium Research Institute, 7700 Sandholdt Rd., Moss Landing, CA 95039, USA

\‡ Corresponding author: Alejandro Damian-Serrano, email: alejandro.damianserrano@yale.edu

## Abstract {-}

Predators have evolved dedicated body parts to capture and subdue prey. As different predators specialize on distinct prey taxa, their tools for prey capture diverge into a variety of adaptive forms. Studying the evolution of predation is facilitated by a predator clade with structures used exclusively for prey capture and with significant morphological variation. Siphonophores, a clade of colonial cnidarians, satisfy these criteria particularly well, capturing prey with their tentilla (tentacle side branches). Earlier work has shown that extant siphonophore diets correlate with the different morphologies and sizes of their tentilla and nematocysts. We hypothesize that evolutionary specialization on different prey types has driven the phenotypic evolution of these characters. To test this hypothesis, we: (1) measured multiple morphological traits from fixed siphonophore specimens using microscopy and high-speed video techniques, (2) built a phylogenetic tree of 45 species, and (3) analyzed the evolutionary associations between siphonophore nematocyst characters and prey type data from the literature. Our results show that siphonophore tentillum structure has strong evolutionary associations with prey type and size specialization, and suggest that shifts between prey-type specializations are linked to shifts in tentillum and nematocyst size and shape. We found that predatory specialists can evolve into generalists, and that specialists on one prey type have directly evolved into specialists on other prey types. In addition, we generated hypotheses about the diets of understudied siphonophore species based on these characters. Thus, the evolutionary history of tentilla shows that siphonophores are an example of ecological niche diversification via morphological innovation and evolution. This study contributes to understanding how morphological evolution has shaped present-day oceanic food webs.

## Keywords {-}

Siphonophores, tentilla, nematocysts, predation, specialization, character evolution

##   {-}

  Most animal predators have characteristic biological tools that they use to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures tend to be finely attuned to their ability to successfully capture specific prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.* avoidance, escape, protective barriers) leads to modifications that can counter those defenses The more specialized the diet of a predator is, the more specialized its tools need to be to meet the specific challenges posed by the prey. Understanding the relationships between predatory specializations and morphological specializations is necessary to contextualize the phenotypic diversity of predators, and to quantify the importance of ecological diversification in generating this diversity.

  Siphonophores (Cnidaria : Hydrozoa) are a clade of organisms bearing modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). These present a significant morphological variation across species [@mapstone2014global] (Fig. \@ref(figure2)), which makes them an ideal system to study the relationships between functional traits and prey specialization. A siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)), each with a single tentacle, which branches into several tentilla carrying the functional cnidocytes (specialized neural cells carrying nematocysts, the stinging capsules). Unlike most other cnidarians, siphonophores carry their tentacle nematocysts in extremely complex and organized batteries [@skaer1988formation] built into their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent]. Siphonophores bear four major nematocyst types in their tentacles and tentilla. The largest type, heteronemes, have open-tip tubules characterized by bearing a distinctly wider spiny shaft at the proximal end of the everted tubule. These are typically found flanking the proximal end of the cnidoband. The most abundant type, haplonemes, have no distinct shaft, but similarly to heteronemes, their tubules have open tips and can be found in the cnidoband. Both heteronemes and haplonemes bear short spines along the tubule and can be toxic and penetrate the surface of some prey types. In the terminal filament, siphonophores bear two other types of nematocysts, characterized by their adhesive function, closed tip tubules, and lack of spines on the tubule. These are the desmonemes (a type of adhesive coiled-tubule spironeme), and rhopalonemes (a siphonophore-exclusive nematocyst type with wide tubules).

  Many siphonophore species inhabit the deep pelagic ocean, which spans from ~200m to the oceanic seafloor. This habitat has fairly homogeneous physical conditions and stable zooplankton abundances and composition [@robison2004deep]. With a relatively predictable prey availability, ecological theory would predict evolution to drive coexisting siphonophore lineages towards specialization, increasing their feeding efficiencies and reducing interspecific competition [@simpson1944tempo;@hardin1960competitive;@hutchinson1961paradox]. If this prediction holds true, we expect the prey capture apparatus morphologies of siphonophores to diversify with the evolution of increased specialization on a variety of prey types in different siphonophore lineages.

Specialization is often thought to be an evolutionary ‘dead end’, meaning that specialized lineages are unlikely to evolve into generalists or to shift the resource for which they are specialized [@futuyma1988evolution]. However, recent studies have found that interspecific competition can favor the evolution of resource generalism [@stireman2005evolution;@johnson2009competition] and resource switching [@hoberg2008macroevolutionary]. Here we examine three alternative hypotheses on siphonophore trophic specialization: (1) predatory specialists evolved from generalist ancestors; (2) predatory specialists evolved from ancestral predators which specialized on a different resource, switching their primary prey type; and (3) predatory generalists evolved from specialist ancestors.

  The study of siphonophore tentilla and diets has been limited in the past due to the inaccessibility of their oceanic habitat and the difficulties associated with the collection of fragile siphonophores. Thus, the morphological diversity of tentilla has only been characterized for a few taxa, and their evolutionary history remains largely unexplored. Contemporary underwater sampling technology provides an unprecedented opportunity to explore the trophic ecology [@choy2017deep] and functional morphology [@costello2015multi] of siphonophores. In addition, well-supported phylogenies based on molecular data are now available for these organisms [@munro2018improved]. These advances allow for the examination of relationships between modern siphonophore form, function, and ecology, as well as reconstructing their evolutionary history.

  The few pioneering studies that have addressed the relationships between tentilla and diet suggest that siphonophores are a robust system for the study of predatory specialization via morphological diversification. [@purcell1984functions] and [@purcell1988correlation] showed clear relationships between diet, tentillum, and nematocyst characters in co-occurring epipelagic siphonophores. These correlations, while studied for a small subset of extant epipelagic siphonophore species, might be generalizable to all siphonophores. We hypothesize that these relationships reflect correlated evolution between prey selection and tentillum (and nematocyst) traits. Furthermore, we hypothesize that with an extensive characterization of tentilla morphology, we can generate hypotheses about the diets of understudied siphonophore species.

In addition, our study design allows us to address other interesting questions about the morphology and evolution of these unique structures. In particular, we aim to address the evolutionary origins of giant tentilla, the character integration of tentilla, the evolution of the extreme shapes of siphonophore nematocysts [@thomason1988allometry], and the mechanical implications of tentillum morphologies on cnidoband discharge.

  In this study, we characterize the morphological diversity of tentilla and their nematocysts across a broad variety of shallow and deep sea siphonophore species using modern imaging technologies, we expand the phylogenetic tree of siphonophores by combining a broad taxon sampling of ribosomal gene sequences with a transcriptome-based backbone tree, and we explore the evolutionary histories and correlations among diet, tentillum, and nematocyst characters.

![\label{figure1} Siphonophore anatomy. A - *Nanomia* sp. siphonophore colony (photo by Catriona Munro). B,C - Illustration of a *Nanomia* colony, gastrozooid, and tentacle (by Freya Goetz). D - *Nanomia* sp. Tentillum illustration and main parts. E - Differential interference contrast micrograph of the tentillum illustrated in D. F - Nematocyst types (illustration reproduced with permission from Mapstone 2014), hypothesized homologies, and locations in the tentillum. Undischarged to the left, discharged to the right.](Figures_final/figure1.jpg)

![\label{figure2} Tentillum diversity plate. The illustrations delineate the pedicle (green), involucrum (blue), cnidoband (orange), elastic strands (pink), terminal structures (yellow). Heteroneme nematocysts (stenoteles in C,E,F,G and mastigophores in H,I) are depicted in red for some species. A - *Erenna laciniata*, 10x. B - *Lychnagalma utricularia*, 10x. C - *Agalma elegans*, 10x. D - *Resomia ornicephala*, 10x. E - *Frillagalma vityazi*, 20x. F - *Bargmannia amoena*, 10x. G - *Cordagalma* sp., reproduced from Carré 1968. H - *Lilyopsis fluoracantha*, 20x. I - *Abylopsis tetragona*, 20x.](Figures_final/figure2.jpg)

## Methods {-}

*Tentillum morphology* – The morphological work was carried out on siphonophore specimens fixed in 4% formalin from the Yale Peabody Museum Invertebrate Zoology (YPM-IZ) collection (accession numbers in Appendix 1). These specimens were collected intact across many years of fieldwork expeditions, using  blue-water diving [@haddock2005scientific], remotely operated vehicles (ROVs), and human-operated submersibles. Tentacles were dissected from non-larval gastrozooids, sequentially dehydrated into 100% ethanol, cleared in methyl salicylate, and mounted onto slides with Canada Balsam or Permount mounting media. The slides were imaged as tiled z-stacks using differential interference contrast (DIC) on an automated stage at YPM-IZ (with the assistance of Daniel Drew and Eric Lazo-Wasem) and with laser point confocal microscopy using a 488 nm Argon laser that excited autofluorescence in the tissues. Thirty characters (defined in Appendix 2) were measured using Fiji [@collins2007imagej;@schindelin2012fiji]. We did not measure the lengths of contractile structures (terminal filaments, pedicles, gastrozooids, and tentacles), since they are too variable to quantify. We measured at least one specimen for 96 different species (Appendix 3, Fig. \@ref(figure3)). Of these, we selected 38 focal species across clades based on specimen availability and phylogenetic representation. Three to five tentacle specimens from each one of these selected species were measured to capture intraspecific variation.

In order to observe the discharge behavior of different tentilla, we recorded high speed footage (1000-3000 fps) of tentillum and nematocyst discharge by live siphonophore specimens (26 species) using a Phantom Miro 320S camera mounted on a stereoscopic microscope. We mechanically elicited tentillum and nematocyst discharge using a fine metallic pin. We used the Phantom PCC software to analyze the footage. For the 10 species recorded, we measured total cnidoband discharge time (ms), heteroneme filament length (µm), and discharge speeds (mm/s) for cnidoband, heteronemes, haplonemes, and heteroneme shafts when possible (data in Appendix 4).

*Siphonophore phylogeny* – The phylogenetic analysis included 55 siphonophore species and 6 outgroup cnidarian species (*Clytia hemisphaerica*, *Hydra circumcincta*, *Ectopleura dumortieri*, *Porpita porpita*, *Velella velella*, *Staurocladia wellingtoni*). The gene sequences we used in this study are available online (accession numbers in Appendix 5). Some of the sequences we used were accessioned in [@dunn2005molecular], and others we extracted from the transcriptomes in [@munro2018improved]. Two new 16S sequences for *Frillagalma vityazi* (MK958598) and *Thermopalia* sp. (MK958599) sequenced by Lynne Christianson were included and accessioned to NCBI. We aligned these sequences using MAFFT [@katoh2002mafft] (alignments available in Dryad). We inferred a Maximum Likelihood (ML) phylogeny (Appendix 6) from 16S and 18S ribosomal rRNA genes using IQTree [@nguyen2014iq] with 1000 bootstrap replicates (iqtree -s alignment.fa -nt AUTO -bb 1000). We used ModelFinder [@kalyaanamoorthy2017modelfinder] implemented in IQTree v1.5.5. to assess relative model fit. ModelFinder selected GTR+R4 for having the lowest Bayesian Information Criterion score. Additionally, we inferred a Bayesian tree with each gene as an independent partition in RevBayes [@hohna2016revbayes] (Appendix 7 and 9), which was topologically congruent with the unconstrained ML tree. The *alpha* priors were selected to minimize prior load in site variation.

Given the broader sequence sampling of the transcriptome phylogeny, we ran constrained inferences (using both ML and Bayesian timetree approaches, which produced fully congruent topologies (Appendix 6 and 7)) after fixing the 5 nodes that were incongruent with the topology of the consensus tree in [@munro2018improved]. This topology was then used to inform a Bayesian relaxed molecular clock time-tree in RevBayes, using a birth-death process (sampling probability calculated from the known number of described siphonophore species) to generate ultrametric branch lengths (Appendix 8). Scripts available in Appendix 9.

*Feeding ecology* – We extracted categorical diet data for different siphonophore species from published sources, including seminal papers [@biggs1977field;@purcell1981dietary;@purcell1984functions;@Mackie:1987uy;@pugh1988two;@bardi2007taxonomic;@andersen1981redescription], and ROV observation data [@choy2017deep;@hissmann2005situ] with the assistance of Elizabeth Hetherington and C. Anela Choy (Appendix 10). We removed the gelatinous prey observations for *Praya dubia* eating a ctenophore and a hydromedusa, and for *Nanomia* sp. eating *Aegina*, since we believe these are rare events that have a much larger probability of being detected by ROV methods than their usual prey, and it is not clear whether the medusae were attempting to prey upon the siphonophores. Personal observations on feeding (from SHDH, CAC, and Philip Pugh) were also included for *Resomia ornicephala*, *Lychnagalma utricularia*, *Bargmannia amoena*, *Erenna richardi*, *Erenna laciniata*, *Erenna sirena*, and *Apolemia rubriversa*. In order to detect coarse-level patterns in the feeding habits, the data were merged into feeding guilds. The feeding guilds described here are: small-crustacean specialist (feeding mainly on copepods and ostracods), large crustacean specialist (feeding on large decapods, mysids, or krill), fish specialist (feeding mainly on actinopterygian larvae, juveniles, or adults), gelatinous specialist (feeding mainly on other siphonophores, medusae, ctenophores, salps, and/or doliolids), and generalist (feeding on a combination of the aforementioned taxa, without favoring any one prey group). These were selected to minimize the number of categories while keeping the most different types of prey separate. We extracted copepod prey length data from [@purcell1984functions]. To calculate specific prey selectivities, we extracted quantitative diet and zooplankton composition data from [@purcell1981dietary], matched each diet assessment to each prey field quantification by site, calculated Ivlev’s electivity indices [@jacobs1974quantitative], and averaged those by species (Appendix 11).

*Statistical analyses* – For subsequent comparative analyses, we removed species present in the tree but not represented in the morphology data, and *vice versa*. Although we measured specimens labeled as *Nanomia bijuga* and *Nanomia cara*, we are not confident in some of the species-level identifications, and some specimens were missing diagnostic zooids. Thus, we decided to collapse these into a single taxonomic concept (*Nanomia* sp.). All *Nanomia* sp. observations were matched to the phylogenetic position of *Nanomia bijuga* in the tree. We carried out all phylogenetic comparative statistical analyses in the programming environment R [@team2017r], using the Bayesian ultrametric species tree (Fig. \@ref(figure4)), and incorporating intraspecific variation estimated from the specimen data as standard error whenever the analysis tool allowed it (Appendix 3). R scripts available in Dryad. For each character (or character pair) analyzed, we removed species with missing data and reported the number of taxa included. We tested each character for normality using the Shapiro-Wilk test [@shapiro1965analysis], and log-transformed those that were non-normal.

We fitted different models generating the observed data distribution given the phylogeny for each continuous character using the function fitContinuous in the R package *geiger* [@harmon2007geiger]. The models compared were the white noise (WN; non-phylogenetic model that assumes all values come from a single normal distribution with no covariance structure among species), the Brownian Motion (BM) model of neutral divergent evolution [@martins1996phylogenies], the Early Burst (EB) model of decreasing rate of evolutionary change [@harmon2010early], and the Ornstein-Uhlenbeck (OU) model of stabilizing selection around a fitted optimum state [@uhlenbeck1930theory;@butler2004phylogenetic]. We then ranked the models in order of increasing parametric complexity (WN,BM,EB,OU), and compared the corrected Akaike Information Criterion (AICc) support scores [@sugiura1978further] to the lowest (best) score, using a cutoff of 2 units to determine significantly better support. When the best fitting model was not significantly better than a less complex alternative, we selected the least complex model (Appendix 12). We calculated model adequacy scores using the R package *arbutus* [@pennell2015model] (Appendix 13). We calculated phylogenetic signal in each of the measured characters using Blomberg’s K [@blomberg2003testing] (Appendix 12). We reconstructed ancestral states using Maximum Likelihood (R phytools::anc.ML [@revell2012phytools]), and stochastic character mapping (R phytools::make.simmap) for categorical characters. R scripts available in Dryad.

In order to study the evolution of predatory specialization, we reconstructed components of the diet and prey selectivity on the phylogeny using ML (R phytools::anc.ML). To identify evolutionary associations of diet with tentillum and nematocyst characters, we compared the performance of a neutral evolution model to that of a diet-driven directional selection model. First, we collapsed the diet data into the five feeding guilds mentioned above (fish specialist, small crustacean specialist, large crustacean specialist, gelatinous specialist, generalist), based on which prey types they were observed consuming most frequently. Then, we reconstructed the feeding guild ancestral states using the ML function ace (package ape [@paradis2019package]), removing tips with no feeding data. The ML reconstruction was congruent with the consensus stochastic character mapping (Appendix 18). Then, using the package *OUwie* [@beaulieu2012ouwie], we fitted an OU model with multiple optima and rates of evolution matched to the reconstructed ancestral diet regimes, a single optimum OU model, and a BM null model, inspired by the analyses in [@cressler2015detecting]. Finally, we compared their AICc support values to select the best fitting model (Appendix 14).

To model the evolutionary associations between individual tentillum and nematocyst characters and the ability to capture particular prey types in the diet, we ran a series of phylogenetic generalized linear models (R phylolm::phyloglm) (Appendix 17). In addition, we ran a series of comparative analyses to address hypotheses of diet-tentillum relationships posed in the literature. To test for correlated evolution among binary characters, we used Pagel’s test [@pagel1994detecting]. To characterize and evaluate the relationship between continuous characters, we used phylogenetic generalized least squares regressions (PGLS) [@grafen1989phylogenetic]. To compare the evolution of continuous characters with categorical aspects of the diet, we carried out a phylogenetic logistic regression (R nlme::gls using the 'corBrownian' function for the argument ‘correlation’).

@@

In order to study correlations between the rates of evolution between different characters, we fitted a set of evolutionary variance covariance matrices [@revell2009phylogenetic] (R phytools::evol.vcv). When fitting all covariance terms simultaneously (Appendix 19.1-20.3), we selected the largest set of characters that would allow the analysis to run without computational singularities. This excluded many of the morphometric characters which are linearly dependent on other characters. Since the functions do not tolerate missing data, we ran the analyses in two ways: One including all taxa but transforming absent states to zeroes, and another removing the taxa with absent states. To test whether phenotypic integration changes across selective regimes determined by the reconstructed feeding guilds, we carried out character-pairwise variance covariance analysis comparing alternative models (R phytools::evolvcv.lite), including those where correlations are the same across the whole tree and models where correlations differ between selective regimes. These analyses could only be carried out on the subset of taxa for which diet data is available, and only among character pairs that are not computationally singular for that taxonomic subset. Finally, we compared regime-specific variance covariance matrices to the general matrix and to their preceding regime matrix to identify the changes in character dependence unique to each regime (see Appendix 19). Gelatinous specialist correlations could only be estimated for a small subset of characters present in *Apolemia*, and should be interpreted with care.

To test how many times extreme nematocyst morphologies evolved, we reconstructed the ancestral states of *log*(length/width) of the different cnidoband nematocyst types, and identified the branches with the greatest shifts. In addition to characterizing the shifts in the state values of haploneme and heteroneme elongation, we identified and located regime shifts for the rate of evolution using a Bayesian Analysis of Macroevolutionary Mixtures (BAMM) [@rabosky2014bamm] (Appendix 16).

## Results {-}

@@

*Phylogeny* – Only 5 nodes (blue dots in Figure \@ref(figure4)) in the unconstrained inference were incongruent with the [@munro2018improved] transcriptome tree, and these were constrained during estimation of the 18S+16S tree. The topology of the constrained tree presented here (Fig. \@ref(figure4)) is congruent with the resolved nodes in [@dunn2005molecular] and [@munro2018improved].

![\label{figure4} Bayesian time-tree built from 18S + 16S concatenated sequences. Branch lengths estimated using relaxed molecular clock. Species names in red indicate replicated representation in the morphology data. Species marked with an asterisk were recorded using high speed video. Nodes labeled with bayesian posteriors (BP). Green circles indicate BP = 1. Blue circles indicate nodes constrained to be congruent with (Munro *et al.* 2018). Tips with black squares indicate the species with transcriptomes used in (Munro *et al.* 2018). Tips with grey squares indicate genus-level correspondence to taxa included in (Munro *et al.* 2018). The main clades are labeled: in black for described taxonomic units, and in grey for operational phylogenetic designations.](Figures_final/figure4.jpg){height=600px}

  We retained the clade nomenclature defined in [@dunn2005molecular] and [@munro2018improved], such as Codonophora to indicate the sister group to Cystonectae, Euphysonectae to indicate the sister group to Calycophorae, Clade A and B to indicate the two main lineages within Euphysonectae. In addition, we define two new clades within Codonophora (Fig. \@ref(figure4)): Eucladophora as the clade containing *Agalma elegans* and all taxa that are more closely related to it than to *Apolemia lanosa*, and Tendiculophora as the clade containing *Agalma elegans* and all taxa more closely related to it than to *Bargmannia elongata*. Eucladophora is characterized by bearing spatially differentiated tentilla with proximal heteronemes and a narrower terminal filament region. The etymology derives from the Greek *eu*+*kládos*+*phóros* for “true branch bearers”. Tendiculophora are characterized by bearing rhopalonemes and desmonemes in the terminal filament, having a pair of elastic strands, and developing proximally detachable cnidobands. The etymology of this clade is derived from the Latin *tendicula* for “snare or noose” and the Greek *phóros* for "carriers".

*Evolutionary dynamics between diet and tentillum morphology* – Reconstructions of feeding guilds shows that generalism is not likely to be ancestral, and it appears to have evolved at least two times independently (Fig. \@ref(figure5)). Large crustacean specialists evolve into generalists twice independently, supporting hypothesis 3. Feeding guild specializations have shifted from an alternative ancestral state at least five times, supporting hypothesis 2. Copepod specialization and fish specialization evolved twice, and ostracod specialization evolved at least once.

The OUwie model comparison shows that out of 30 characters, 10 show significantly stronger support for the diet-driven multi-optima multi-rate OU model (Appendix 14). These characters include terminal filament nematocyst size and shape, involucrum length, elastic strand width, and heteroneme number. Most of these characters are found exclusively in Tendiculophora, thus this may reflect processes that could be unique to this subtree. Five characters including cnidoband length, cnidoband shape, and haploneme length show maximal support for a diet-driven single-optimum OU model. The remaining 15 characters support BM (or OU with marginal AICc difference with BM).

![\label{figure5} Left - Subset phylogeny showing the mapped feeding guild regimes that were used to inform the *OUwie* analyses. Right - Grid showing the prey items consumed from which the feeding guild categories were derived. Diet data were obtained from the literature review in Appendix 10.](Figures_final/figure5.jpg)

  Phylogenetic logistic regressions identified evolutionary associations between individual characters and the presence of particular prey types in the diet (Fig. \@ref(figure5), right). Shifts toward ostracod presence in diet correlated with reductions in pedicle width and total haploneme volume. Shifts to copepod presence in the diet were associated with reductions in haploneme width, cnidoband length and width, total haploneme and heteroneme volumes, and tentacle and pedicle widths. Consistently, transitions to decapod presence in the diet correlated with more coiled cnidobands (Appendix 17).

  Phylogenetic regressions of continuous characters against prey selectivity data produced additional insights. Fish selectivity is associated with increased number of heteronemes per tentillum, increased roundness of nematocysts (desmonemes and haplonemes), larger heteronemes, reduced heteroneme/cnidoband length ratios, smaller rhopalonemes, lower haploneme SA/V ratios, and increased size of the cnidoband, elastic strand, pedicle and tentacle widths. Decapod-selective diets were associated with increasing cnidoband size and coiledness, haploneme row number, elastic strand width, and heteroneme number. Copepod-selective diets evolved in association with smaller heteroneme and total nematocyst volumes, smaller cnidobands, rounder rhopalonemes, elongated heteronemes, narrower haplonemes with higher SA/V ratios, and smaller heteronemes, tentacles, pedicles and elastic strands. Selectivity for ostracods was associated with reductions in size and number of heteroneme nematocysts, reductions in cnidoband size, number of haploneme rows, heteroneme number, and cnidoband coiledness. Heteroneme length and elongation also correlated negatively with chaetognath selectivity.

  When some of the diet-morphology associations reported in the literature [@purcell1984functions;@purcell1988correlation] were tested for correlated evolution (Table 1), we found that most were consistent with an evolutionary explanation except the relationship between terminal filament nematocysts (rhopalonemes and desmonemes) and crustaceans in the diet. The latter is likely a product of the larger species richness of crustacean-eating species with terminal filament nematocysts, rather than simultaneous evolutionary gains.

\

Table 1. Tests of correlated evolution between morphological characters and aspects of the diet found correlated in the literature.

```{r tab1, echo=FALSE, out.width='100%'}
knitr::include_graphics('Final_tables/table1.png')
```
\

@@

*Phenotypic integration of the tentillum* – The quantitative characters we measured from tentilla and their nematocysts are highly correlated. The results indicate that the dimensionality of tentillum morphology is low, that many traits are associated with size, but that nematocyst arrangement and shape are independent of it. Of the phylogenetic correlations (Fig. \@ref(figure8)a, lower triangle), 81.3% were positive and 18.7% were negative, while of the ordinary correlations (Fig. \@ref(figure8)a, upper triangle) 74.6% were positive and 25.4% were negative. Half (49.9%) of phylogenetic correlations were >0.5, while only 3.6% are < -0.5. Similarly, of the across-species correlations, 49.1% were >0.5 and only 1.5% were < -0.5. We found that 13.9% of character pairs had opposing phylogenetic and ordinary correlation coefficients. Just 4% have negative phylogenetic and positive ordinary correlations (such as rhopaloneme elongation ~ heteroneme-to-cnidoband length ratio and haploneme elongation, or haploneme elongation ~ heteroneme number), and only 9.9% of character pairs had positive phylogenetic correlation yet negative ordinary correlation (such as heteroneme elongation ~ cnidoband convolution and involucrum length, or rhopaloneme elongation with cnidoband length). These disparities can be caused by Simpson’s paradox [@blyth1972simpson]: the reversal of the sign of a relationship when a third variable (or a phylogenetic topology [@uyeda2018rethinking]) is considered. However, no character pair had correlation coefficient differences larger than 0.64 between ordinary and phylogenetic correlations (heteroneme shaft extension ~ rhopaloneme elongation has a Pearson’s correlation of 0.10 and a phylogenetic correlation of  -0.54). Rhopaloneme elongation shows the most incongruencies between phylogenetic and ordinary correlations with other characters.

The variance covariance matrices (Appendix 19.1-20.2) are congruent with the abundant positive correlations observed among simple measurement characters in Fig. \@ref(figure8)a. However, this analysis reveals more clearly the diagonal blocks that constitute the evolutionary modules, such as the heteroneme block, the terminal filament nematocyst block, and the cnidoband-pedicle-tentacle block. These results were not very sensitive to transformation of inapplicable states and taxon sampling. When we compared the rate covariance terms between characters across the different feeding guild regimes (Appendix 19.4), we found that half (48%) of the character pairs presented distinct correlation coefficients across different regimes, indicating that the mode of phenotypic integration may also shift with trophic niche. When contrasting the regime-specific rate correlation matrices to the whole-tree matrix, we were able to identify the character dependencies that are unique to each predatory niche (Appendix 19.6).

Under the majority of SIMMAP outcomes, large crustaceans specialists are the first regime to appear, and other regimes evolve in a shift from this ancestral specialization. Compared to the rate correlation matrix estimated over the whole tree, large crustacean specialists present strong negative correlations between haploneme elongation and heteroneme size, and between rhopaloneme elongation and tentillum size, as well as with involucrum length. With the appearance of generalists (*Forskalia* and the *Agalma-Athorybia* clade), terminal filament nematocyst (desmonemes and rhopalonemes) sizes became negatively correlated with the sizes of most characters, meaning that as some tentilla became larger, their individual terminal nematocysts became smaller, observed to the extreme in *Agalma*. In addition, heteroneme and rhopaloneme elongation became positively correlated with cnidoband size. When large crustacean specialists switched to small crustacean prey in *Cordagalma* and calycophorans, haploneme size became inversely correlated with heteroneme elongation, which in turn developed a strong positive relationship with tentillum size. In other words, as tentilla get smaller in this group, heteronemes get shorter and haplonemes get larger. The extremes of this gradient can be seen in *Cordagalma* and *Hippopodius*. With the evolution of fish prey specialization in cystonects and within Clade B, haploneme elongation became negatively correlated with heteroneme elongation (signal driven by Clade B, since cystonects lack tentacular heteronemes), and the surface area to volume ratio of haploneme nematocysts switched from a strong negative relationship with cnidoband size (found in every other regime) to a positively correlation. Gelatinous specialization, albeit appearing only once in our tree, also carries a unique signature in character rate correlation shifts, with an increase in the strength of the correlation between heteroneme shape and shaft width, consistent with the appearance of birrhopaloid nematocysts with swollen shafts that are likely effective at anchoring gelatinous tissue (see reference to Narcomedusae nematocysts in [@purcell1988correlation]).

![\label{figure9} Phylomorphospace of the simple continuous characters principal components, excluding ratios and composite characters. A. Variance explained by each variable in the PC1-PC2 plane. Axis labels include the phylogenetic signal (K) for each component and p-value. B. Phylogenetic relationships between the species points distributed in that same space.](Figures_final/figure9.jpg){height=600px}

  In the non-phylogenetic PCA morphospace using only simple characters (Fig. \@ref(figure9)), PC1 (aligned with tentillum and tentacle size) explained 69.3% of the variation in the tentillum morphospace, whereas PC2 (aligned with heteroneme length, heteroneme number, and haploneme arrangement) explained 13.5%. In a phylogenetic PCA, 63% of the evolutionary variation in the morphospace is explained by PC1 (aligned with shifts in tentillum size), while 18% is explained by PC2 (aligned with shifts in heteroneme number and involucrum length).

*Evolution of nematocyst shape* – Haploneme nematocyst evolution has been mainly driven by a single large shift towards elongation in Tendiculophora, which contains the majority of described siphonophore species other than Cystonects, *Apolemia*, and Pyrostephidae. There is one secondary return to more oval, less elongated haplonemes in *Erenna*, but it does not reach the sphericity present in Cystonectae or Pyrostephidae (Fig. \@ref(figure10). Heteroneme evolution presents a less discrete evolutionary history, where Tendiculophora evolved more elongate heteronemes, but the difference between theirs and other siphonophores is much smaller than the variation in shape within Tendiculophora, bearing no phylogenetic signal. In this clade, evolution of heteroneme shape has diverged in both directions, and there is no correlation with haploneme shape (Fig. \@ref(figure10)), which has remained fairly constant (elongation between 1.5 and 2.5).

![\label{figure10} Phylomorphospace showing haploneme and heteroneme elongation (log scaled). Orange area delimits rod-shaped haplonemes, blue area covers oval and round shaped haplonemes. Smaller dots and lines represent phylogenetic relationships and ancestral states of internal nodes under BM. Species nodes in red were manually added to the plot. Cystonects have no tentacle heteronemes and are projected onto the haploneme axis. Apolemiids have no tentacle haplonemes and are projected onto the heteroneme axis. Colored branches and nodes correspond to BAMM regimes of accelerated haploneme shape (green) and heteroneme shape (violet) evolution.](Figures_final/figure10.jpg)

  Haploneme and heteroneme shape share 21% of their variance across extant values, and 53% of variance in their shifts along the branches of the phylogeny. However, much of this correlation is due to the contrast between Pyrostephidae and their sister group Tendiculophora (Fig. \@ref(figure4)). BAMM identified a regime shift in heteroneme shape evolution on the branches leading to *Agalma* and *Athorybia*. For the rates of haploneme shape evolution, BAMM identified two main independent regime shifts (Fig. \@ref(figure10)): one in the branch leading to Codonophora (anisorhizas diverging from cystonects’ spherical isorhizas), and one in the branch leading to Clade B physonects. Clade B includes *Erenna*, *Stephanomia*, *Marrus*, and rhodaliids. Most of these taxa have rod-shaped anisorhizas, but *Erenna* has oval ones). No clear regime shift patterns were identified in the evolution of desmoneme and rhopaloneme shape.

*Functional morphology of tentillum and nematocyst discharge* –  Tentillum and nematocyst discharge high speed measurements are available in Appendix 4. While the sample sizes of these measurements were insufficient to draw reliable statistical results at a phylogenetic level, we did observe patterns that may be relevant to their functional morphology. For example, cnidoband length is strongly correlated with discharge speed (p value = 0.0002). This is probably the sole driver of the considerable difference between euphysonect and calycophoran tentilla discharge speeds (average discharge speeds: 225.0mm/s and 41.8mm/s respectively; t-test p value = 0.011), since the euphysonects have larger tentilla than the calycophorans among the species recorded.

  We also observed that calycophoran haploneme tubules fire faster than those of euphysonects (T-test p value = 0.001). Haploneme nematocysts discharge 2.8x faster than heteroneme nematocysts (T-test p value = 0.0012). Finally, we observed that the stenoteles of the Euphysonectae discharge a helical filament that “drills” itself through the medium it penetrates as it everts.

>>>>

## Supplementary Methods {-}

@@

## Supplementary Results {-}

@@

@@

*Phenotypic integration of the tentillum* – Of the phylogenetic correlations (Fig. \@ref(figure8)a, lower triangle), 81.3% were positive and 18.7% were negative, while of the ordinary correlations (Fig. \@ref(figure8)a, upper triangle) 74.6% were positive and 25.4% were negative. Half (49.9%) of phylogenetic correlations were >0.5, while only 3.6% are < -0.5. Similarly, of the across-species correlations, 49.1% were >0.5 and only 1.5% were < -0.5. We found that 13.9% of character pairs had opposing phylogenetic and ordinary correlation coefficients. Just 4% have negative phylogenetic and positive ordinary correlations (such as rhopaloneme elongation ~ heteroneme-to-cnidoband length ratio and haploneme elongation, or haploneme elongation ~ heteroneme number), and only 9.9% of character pairs had positive phylogenetic correlation yet negative ordinary correlation (such as heteroneme elongation ~ cnidoband convolution and involucrum length, or rhopaloneme elongation with cnidoband length). These disparities can be caused by Simpson’s paradox [@blyth1972simpson]: the reversal of the sign of a relationship when a third variable (or a phylogenetic topology [@uyeda2018rethinking]) is considered. However, no character pair had correlation coefficient differences larger than 0.64 between ordinary and phylogenetic correlations (heteroneme shaft extension ~ rhopaloneme elongation has a Pearson’s correlation of 0.10 and a phylogenetic correlation of  -0.54). Rhopaloneme elongation shows the most incongruencies between phylogenetic and ordinary correlations with other characters.
